---
title: "Signal Detection Theory"
subtitle: |
  Analysing the 2-point touch discrimination data
author: "Murillo Pagnotta"
date: "2025-09-23"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_width: 10
    code_folding: show
    number_sections: true
---


# Preface

This document is based on:

Zloteanu, M., & Vuorre, M. (2024). A Tutorial for Deception Detection Analysis or: How I Learned to Stop Aggregating Veracity Judgments and Embraced Signal Detection Theory Mixed Models. Journal of Nonverbal Behavior, 48(1), 161–185. https://doi.org/10.1007/s10919-024-00456-x


# Set up
```{r setup and packages, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r}
library(tidyverse)
```


# Data
```{r}
# Make sure the data file and analysis file are in the same folder

data = read_delim("full_data.csv", delim = ";")

# Anonomysing data
anon_col <- function(x, prefix) {
  u <- unique(x)
  new <- paste0(prefix, formatC(sample(length(u)), width = nchar(length(u)), flag = "0"))
  new[match(x, u)]
}

data = data %>%
  mutate(
    group_id = anon_col(group_id, "G"),
    experimenter_id = anon_col(experimenter_id, "E"),
    participant_id = anon_col(participant_id, "P")
  )

data %>%
  mutate(
    participant_id = factor(participant_id),
    bodypart = factor(bodypart),
    stim_type = factor(stim_type),
    distance = factor(distance)
  ) 

# check NAs
# colSums(is.na(d))

glimpse(data)
```

How many participants per body_part?
```{r}
data %>% group_by(bodypart, participant_id) %>%
  slice_head(n = 1) %>%
  group_by(bodypart) %>%
  count()
```

How many trials per participant?
```{r}
data %>% count(participant_id)
```


# Non-SDT analysis

We first analyse these data with correct / incorrect classifications. Typically, this amounts to aggregating each person's data to proportions correct (in different conditions / groups), and then visualizing and modelling those.

First, we create a new variable encoding response accuracy.

```{r}
data <- data %>% 
  mutate(
    accuracy = case_when( 
      stim_type == "one point" & response == 1 ~ 1,
      stim_type == "two points" & response == 2 ~ 1,
      .default = 0
    ))


```

We then summarize the Accuracy for each body_part and each participant_id to obtain proportion of correct responses.

```{r}
data_acc <- data %>% 
  group_by(bodypart, participant_id) %>%
  summarise(
    accuracy = mean(accuracy)
  )
```

We can then display these proportions correct as a simple histogram by body part.

```{r}
data_acc %>%
  ggplot() +
  geom_histogram(aes(x = accuracy),
                 color = "palevioletred",
                 fill = "rosybrown1"
  ) +
  scale_x_continuous(
    name= "Percent correct",
    limits = c(0, 1)
  ) +
  scale_y_continuous(name = "Count"
  ) +
  facet_wrap(~ bodypart) # facet wrap lets you make a different subplot for a variable of your choice within this plot
```


Run a t-test (separately for each body part) to examine if participants are responding at chance level  (50%) on average

Fingertip
```{r}
data_acc_fingertip = data_acc %>% 
  filter(bodypart == "fingertip") 

t.test(data_acc_fingertip$accuracy, mu = 0.5) 
```
Check the p-value. It is very small. So, we found statistical evidence agaist the null hypothesis and in favour of the alternative hypothesis. In other words, participants were NOT responding at chance level.

Upper arm
```{r}
data_acc_upperArm = data_acc #  ADJUST THE CODE!
```
Check the p-value. Is there evidence that participants were responding above chance level?

We can also summarize the data to proportion correct (Accuracy) for each person at each stim_distance.
Let's also compute the population average.

```{r}
data_acc_dist <- data %>% 
  filter(!distance == "NA") %>% # remove noise trials
  group_by(bodypart, distance, participant_id) %>%
  summarise(
    accuracy = mean(accuracy)
  )

# compute also the population average
data_acc_dist_sum <- data_acc_dist %>% 
  group_by(bodypart, distance) %>%
  summarise(
    pop_accuracy = mean(accuracy)
  )

```


Plot accuracy (y-axis) over distance (x-axis) using geom_point() and geom_line() per participant.

If you can, add a line for the population mean as well
```{r}
ggplot(data = data_acc_dist) # + ??

# Hint: used facet_wrap() to create separate plots for body parts

```


However, this simple measure of accuracy does not differentiate between bias (e.g. to respond "1" over "2") and the ability to discriminate between one-point and two-point stimuli. Signal Detection Theory allows studying bias and sensitivity to two-point touches separately.

# Signal Detection Theory

In practice, the parameters of a basic SDT model can be calculated from two proportions. To calculate those proportions, we first classify each trial into four different categories:

**Hits** are trials in which a two-point stimulus is presented, and the participant correctly responds "two".

**Correct rejections** are trials in which a one-point stimulus is presented, and the participant correctly responds "1".

**False alarms** are trials in which a one-point stimulus is presented, but the participant incorrectly responds "2".

**Misses** are trials in which a two-point stimulus is presented, but the participant incorrectly responds "1".

```{r}
sdt_data <- data %>% 
  mutate(
    Type = case_when(
      stim_type == "two points" & response == 2 ~ "hit", # Hit
      stim_type == "one point" & response == 1 ~ "cr",  # Correct rejection
      stim_type == "two points" & response == 1 ~ "miss", # Miss 
      stim_type == "one point" & response == 2 ~ "fa", # False alarm
    )
  )


glimpse(sdt_data)
```

Below is a table illustrating all the classifications' counts in the example data.

```{r}
sdt_data %>% 
  group_by(bodypart) %>%
  count(stim_type, response, accuracy, Type)
```


To calculate the SDT parameters, we must then aggregate the data and compute counts of these four categories for each stim_distance and for each participant. For computations, we also pivot the data to a "wide" format.


```{r}
 sdt_data <- sdt_data %>% 
    # Aggregate per participant x distance
  count(bodypart, participant_id, distance, Type) %>%
  pivot_wider(
    names_from = Type,
    values_from = n,
    values_fill = 0
  )
```



The **hit rate** is then the proportion of "hits" over the sum of "hits" and "misses". The **false alarm rate** is the proportion of "false alarms" over the sum of false alarms and correct rejections. Those rates are then "z-scored" to establish the latent continuum as a standard normal distribution. Sensitivity, often called *d-prime*, is then the distance between the distributions' means. The response criterion is calculated as negative half of the sum of the z-scores. In practice, we calculate all these in R as follows:

```{r}

sdt_data <- sdt_data %>% 
  mutate(
    # hr = hit / (hit + miss),
    hr = (hit + 0.5) / (hit + miss + 1), # with log-linear correction (because we have some zeros) 
    # you can think of the (hit + miss + 1) as (hit + 0.5 + miss + 0.5)
    
    # far = fa / (fa + cr),
    far = (fa + 0.5) / (fa + cr + 1), # with log-linear correction (because we have some zeros)
    
    zhr = qnorm(hr),
    zfa = qnorm(far),
    dprime = zhr - zfa,
    crit = -0.5 * (zhr + zfa)
  )


glimpse(sdt_data)
```



We can then compute basic statistical tests on these person-specific parameters. For example, we can ask whether sensitivity (dprime) was greater than zero.

FINGERTIP
```{r}
sdt_fingertip <- sdt_data %>%
  filter(bodypart == "fingertip")

t.test(sdt_fingertip$dprime) 
```
Check the p-value. Is there evidence for the alternative hypothesis that dprime was greater than zero?


And whether participants show a liberal bias (negative criterion) or a conservative one (positive criterion).

```{r}
t.test(sdt_fingertip$crit) 
```
Check the p-value. Is there evidence for the alternative hypothesis that criterion was liberal or conservative?


UPPER ARM
```{r}
sdt_upperarm <- sdt_data %>%
  filter(bodypart == "upperarm")

t.test(sdt_upperarm$dprime) 
```

And whether participants show a liberal bias (negative criterion) or a conservative one (positive criterion).

```{r}
t.test(sdt_upperarm$crit) 
```

Let's summarise the dataset to plot
```{r}
sdt_sum <- sdt_data %>%
  group_by(bodypart, distance) %>%
  summarise(
    pop_dprime = mean(dprime),
    pop_crit = mean(crit)
  )
```



Plot dprime over distance
```{r}

ggplot(data = sdt_data) # + ??

```



Plot criterion over stim_dist
When criterion is above zero, it is called 'conservative'. In this case, participants are prone to misses (less likely to respond 'two-points').

When criterion is below zero, it is called 'liberal'. In this case, participants are prone to false alarms (more likely to respond 'two-points').

When criterion is zero, participants are unbiased.

```{r}
ggplot(data = sdt_data) +
  geom_point(aes(x = distance , y = crit, colour = participant_id)) +
  geom_line(aes(x = distance , y = crit, colour = participant_id, group = participant_id)) +
  
  # add the population average
  geom_line(data = sdt_sum,aes(x = distance , y = pop_crit, group = NA), colour = "black", size = 1) +
  facet_wrap(~bodypart, scales = "free_x") + 
  
  theme_bw() + 
  theme(legend.position = "none")  

```


# Accuracy as a Generalised Linear Model

Here we model accuracy as a function of distance with a probit GLM (note that only two point trials are included, as the model drops NA rows). The GLM shows us the pattern of change in accuracy, but it cannot tell us why the change occurs, which is what SDT (d' and criterion) adds.


## Fingertip
```{r}
data_fingertip <- data %>% filter(bodypart == "fingertip")
fit1 <- glm(accuracy ~ distance, data = data_fingertip, family = binomial("probit"))

coef(summary(fit1))
```

## Upperarm
```{r}
data_upperarm <- data %>% filter(bodypart == "upperarm")
fit2 <- glm(accuracy ~ distance, data = data_upperarm, family = binomial("probit"))

coef(summary(fit2))
```

Compare the glm with the d-prime and criteria obtained above!
*Answer:*
